---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 16 / 2,
  fig.height = 9 / 2,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

# IxPopDyMod: A framework for Ixodidae Population Dynamics Models

This package is designed to help the user specify, run, and then visualize and analyze the results of Ixodidae (hard-bodied ticks) population dynamics models. Such population dynamics models exist in the literature, but the source code to run them is not always available. We wanted to provide an easy way for these models to be written and shared.

TODO see paper for a full description of the model

## Installation

Install the package from the [GitHub](https://github.com/) with: 
```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("dallenmidd/IxPopDyMod")
```

## Example analysis

We present an example analysis using a simple model configuration pre-loaded into the package. In this example, we will:

1. Vary a parameter in the model
2. Run the model with each new parameter value
3. Calculate the growth rate for each of the model outputs
4. Graph the population over time for each out of the outputs

```{r}
library(IxPopDyMod)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
```

### Vary a parameter in the model
We give a new range of parameter values for number of eggs laid.
```{r}
eggs_laid <- c(800, 1000, 1200)
modified_configs <- vary_param(config_ex_1, from = '__a', to = '__e', 
                               param_name = 'a', values = eggs_laid)
```

### Run the model with each new parameter value
```{r}
outputs <- run_all_configs(modified_configs)
outputs[[1]]
```
The model output is a data frame with columns... Since we ran the model with multiple configurations, we get a list of data frames.

### Calculate growth rate for each of the model outputs
```{r}
sapply(outputs, growth_rate) 
```
The population is stable with 1000 eggs laid, as indicated by the growth rate `1`. The population decreases with 800 eggs laid, and increases with 1200 eggs laid.

### Graph outputs
To see a breakdown of how the population is changing, we graph the population over time of each age group, for each model output. As expected, there is a cycle with a peak in number of eggs, followed by peaks in larvae, nymph and then adult population.

```{r warning = FALSE}
names(outputs) <- c('0800 eggs laid', '1000 eggs laid', '1200 eggs laid')
outputs_stacked <- dplyr::bind_rows(outputs, .id = "id")
outputs_stacked %>%
  graph_population_each_group() +
  facet_wrap(~ id)
```


