---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 16 / 2,
  fig.height = 9 / 2,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

# IxPopDyMod: A framework for Ixodidae Population Dynamics Models

This package is designed to help the user specify, run, and then visualize and analyze the results of Ixodidae (hard-bodied ticks) population dynamics models. Such population dynamics models exist in the literature, but the source code to run them is not always available. We wanted to provide an easy way for these models to be written and shared.

TODO see paper for a full description of the model

## Installation

Install the package from [CRAN](https://cran.r-project.org/web/packages/IxPopDyMod/index.html) with: 
```{r, eval = FALSE}
install.packages("IxPopDyMod")
```

## Examples 

Here we provide a series of examples to help others see how models are specified and better understand the structure of the package. The examples highlight:

1. A very simple example to show basic package commands
1. How tick transitions can be temperature-dependent
1. How to include the host community into a model
1. How to include tick-borne disease infection dynamics into a model
1. How to include host-host density dependent tick mortality

## Simple example

We present an example analysis using a simple model configuration pre-loaded into the package. In this example, we will:

1. Vary a parameter in the model
2. Run the model with each new parameter value
3. Calculate the growth rate for each of the model outputs
4. Graph the population over time for each out of the outputs

If you wish to create a custom model configuration, see `?config()`.

```{r}
library(IxPopDyMod)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
```

### Vary a parameter in the model
We start with `config_ex_1`, a simple model configuration that doesn't consider infection, and that has four life stages: `__e` for egg, `__l` for larvae, `__n` for nymph, and `__a` for adult. We give a new range of parameter values for number of eggs laid.

```{r}
eggs_laid <- c(800, 1000, 1200)
modified_configs <- vary_param(config_ex_1, from = '__a', to = '__e', 
                               param_name = 'a', values = eggs_laid)
```
This gives us a list of three modified model `config`s, which differ only in the number of eggs laid.

### Run the model with each new parameter value
```{r}
outputs <- run_all_configs(modified_configs)
outputs[[1]]
```
The model output is a data frame where the column `day` indicates Julian date, `stage` indicates tick life stage, and `pop` is population size. The remaining columns breakdown the `stage` column into it's constituent parts: the `age` and current `process` of a tick, and whether it is `infected`. Since we ran the model with multiple configurations, we get a list of data frames. Here we inspect only the first.

### Calculate growth rate for each of the model outputs
```{r}
sapply(outputs, growth_rate) 
```
`growth_rate` calculates the multiplicative growth rate for a model output. The population is stable with 1000 eggs laid, as indicated by the growth rate `1`. The population decreases with 800 eggs laid, and increases with 1200 eggs laid.

### Graph outputs
To see a breakdown of how the population is changing, we graph the population over time of each age group, for each model output. As expected, for each output there is a cycle with a peak in number of eggs, followed by peaks in larvae, nymph and then adult population.

```{r warning = FALSE}
names(outputs) <- c('0800 eggs laid', '1000 eggs laid', '1200 eggs laid')
outputs_stacked <- bind_rows(outputs, .id = "id")
outputs_stacked %>%
  graph_population_each_group() +
  facet_wrap(~ id)
```

## Temperature-dependent transitions

In this second example we load in an external `config`. The first step is always to read a `config` file. This is the configuration which specifies the model. Here I read the `config` and then show the tick life-stage transitions.

```{r}
cfg <- read_config('vignettes/temp_example_config/config.yml')
cfg$transitions
```

From the first line, you see that the development from eggs, `__e`, to questing larvae, `q_l`, is an exponential function of temperature. We can see the parameters for this transition:

```{r}
cfg$parameters %>% filter(from == '__e', to == 'q_l')
```

The daily development rate is $0.0000292*temp^{2.27}$.

### Compare two temperature scenarios

Here I highlight how this temperature dependence affects the output of the model. I make a second `config` in which the daily temperature is one degree warmer.

```{r, results = 'hide'}

cfg2 <- cfg
cfg2$weather <- cfg$weather %>% mutate(tmean = tmean + 1)

output1 <- run(cfg)
output2 <- run(cfg2)

output1 <- output1 %>% mutate(temp = 'cold')
output2 <- output2 %>% mutate(temp = 'warm')
```

Finally I compare the outputs for a commonly measured aspect of tick populations, the number of questing nymphs. 

```{r}
output1 %>%
  rbind(output2) %>% 
  filter(stage == 'q_n') %>%
  ggplot(aes(day, pop, col = temp)) +
  geom_line() +
  scale_color_manual(values = c('cold' = 'blue', 'warm' = 'red')) +
  ylab('Questing nymphs')
```

Here you can see nymphs start questing earlier and reach a higher population in the warmer climate. 

## Host community

In the previous example there was no host community explicitly stated and ticks had a constant probabilty of transition between life stages (e.g., from larva to nymph). It is possible to instead model these probabilities based on host community composition. 

```{r}
cfg <- read_config('vignettes/host_example_config/config.yml')
cfg$transitions
```

Here transition from questing larvae, `q_l`, to engroged larvae, `e_l`, depends on the `host_den`, this is how the host community is included in the transition. 

```{r}
cfg$parameters %>% filter(from == 'q.l', to == 'e.l')

```
Here the parameters of `find_n_feed` get different values for each host species. In this case the two parameters are `pref`, which is the larval tick's preference for the three different host species, and `feed_success`, which is the fraction of feeding larvae which successfully feed to completion. 

In this example the temperature and host community are constant through time, but this case change to see how seasonal or year-to-year variation in host density affects tick populations.

### Compare host communities of different densities

I now compare how different host densities affect tick populations. Here I vary the deer density.

```{r, results = 'hide'}
cfg_lowdeerden <- cfg
cfg_highdeerden <- cfg
cfg_lowdeerden$host_comm <- cfg$host_comm %>% 
  mutate(host_den = ifelse(host_spp == 'deer', 0.1, host_den) )

cfg_highdeerden$host_comm <- cfg$host_comm %>% 
  mutate(host_den = ifelse(host_spp == 'deer', 5, host_den) )


output_middeerden <- run(cfg)
output_lowdeerden <- run(cfg_lowdeerden)
output_highdeerden <- run(cfg_highdeerden)

output_middeerden <- output_middeerden %>% mutate(deer_den = 'mid')
output_lowdeerden <- output_lowdeerden %>% mutate(deer_den = 'low')
output_highdeerden <- output_highdeerden %>% mutate(deer_den = 'high')
```

And then use the `graph_population_each_group` function to see how the deer densities affect the tick population.

```{r}
output_lowdeerden %>%
  bind_rows(output_middeerden, output_highdeerden) %>%
  mutate(deer_den = factor(deer_den, levels = c('low', 'mid', 'high'))) %>%
  graph_population_each_group() +
  facet_wrap(~deer_den)

```

## Tick-borne disease infection dynamics

In the examples above I modeled just a tick population without a tick borne disease. Here I give an example of how the package can be used to also include 


So far all examples use transition functions loaded into the package, here we show how to define our own.

```{r}
find_host <- function(x, y, a, pref)
{
  1-(1-a)^sum(x*pref)
}
```



```{r}
cfg <- read_config('vignettes/infect_example_config/config.yml')
cfg$transitions
```
Here we use the middle character of the life-stage key. It is either `i` for infected or `u` for uninfected. We assume no transovarial infection so questing larvae are uninfected. But after they feed, `f_l`, they can either become engorged infected, `eil`, or engorged uninfected, `eul`, larvae. This is based on `infect_fun`, which as above has host species-specific parameters of `pref` and `host_rc` (reservoir competence). 

### Effect of deer density

Deer are important to the blacklegged tick as the main host of adult ticks. As such they are thought to increase tick populations (see above). But deer can also serve as hosts for juvenile tick life stages, and deer are poor reservoirs for *Borrelia burgdorferi*. So increased deer density may also decrease the proportion of bloodmeals juvenile ticks take from more comptenet reservoirs life mice. We use this simple model to illustrate this possiblity. 

```{r}
deer_den <- c(0.1, 0.25, 0.5, 0.75, 1)
results_tib <- tibble(deer = deer_den, nymph_den = 0, nip = 0)

for (i in 1:5)
{
  cfg_mod <- cfg
  cfg_mod$host_comm[1,2] <- deer_den[i]
  out <- run(cfg_mod)
  
  nymph_sum <- out %>%
  filter(stage == 'qin' | stage == 'qun') %>%
  group_by(stage) %>%
  summarise(totpop = sum(pop)) 

  results_tib$nip[i] <- unlist(nymph_sum[1,2] /(nymph_sum[1,2] + nymph_sum[2,2]))
  results_tib$nymph_den[i] <- out %>%
    filter(stage == 'qin' | stage == 'qun') %>%
    summarise(totpop = sum(pop)) %>%
    unlist()
}

results_tib %>%
  ggplot(aes(deer, nip)) +
  geom_point()

results_tib %>%
  ggplot(aes(deer, nymph_den)) +
  geom_point()

```

Here we see that as deer density increases the number of nymphs increases (as the tick population gets bigger). But the nymph infection prevalence (NIP) goes down.
