---
title: "A framework for tick population and infection modeling"
output: html_document
author: 
  - Myles Stokowski
  - Dave Allen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Ticks are obligate, blood-feeding parasites and vectors of important wildlife, livestock and human pathogens (sonenshine2014). It is important to understand tick population dynamics.THis is hard because 

Models for tick populations offer important tools to understand population dynamics. They can be used to understand how tick populations will respond to climate change (ogden2014, winter202, li2016). Models can also be used to test potential tick and tick-borne disease control strategies (mount1997, wang2016). Models can also be used to interrogate theories of tick-borne disease transmission that are hard to address experimentally (ogden2009, levi2016, dunn2013).

There is a long history of modeling tick populations (mount1997, randolph1997, ginsberg1988, haile1987). These models can take a number of different forms ordinary differential equations (maliyoni2017, wallace2019), difference equations (ogden2005), Leslie models (dobson2011, estradapena2014), agent-based (halsey2018) . As with other models there is a tradeoff between model complexity and ability to arrive at analytic results versus the need for simulation. Approximation, such as the next generation matrix, can be used to calcualte the r_0 of either the tick population () or hte tick-borne disease (hartemink2008) for analytic results.

Importance of replicable and reproducible science. How open code fits into this.

Broad philosophy behind our modeling structure. Allows user to easily specify, run, see the results of, and compare tick population models. 

## Package use

```{r load-model}

```

# Specific model with `config`

Required aspects of `config` and then optional aspects. 

# Run, analyze, and graph model

# Compare model output with different parameterization

## Full model example

Here we do the winter tick on moose example. HEre we could use the model to test how climate change affects winter tick populations.

## Conclusions and future directions

## Appendix 1: reproduce Ogden model
